---
title: "Driver mutations and hot spots"
vignette: >
  %\VignetteIndexEntry{quarto vignettes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
    cache: true
from: markdown+emoji
warning: false
fig.width: 9
fig.height: 6
fig.align: "center"
---

# Tutorial: driver mutations and hot spots

In the [previous tutorial](./feature_matrix.html) we assembled a binary feature
matrix using `assemble_genetic_features()`. That approach treats each gene as a
single binary feature: a sample is either mutated or not. For many analyses
that is exactly what you want — but for some genes the *type* of mutation
matters as much as its presence and you will want to separately handle mutations of different varieties or with certain functional annotations.

Consider *TP53*: a truncating mutation, a DNA-contact hotspot substitution
(e.g. R248W), and a structural-domain mutation (e.g. Loop L3) all hit the same
gene but have different biological consequences and could different frequencies across lymphoma subtypes. Similarly, *CREBBP* KAT domain missense mutations have a function distinct from truncating mutations. The canonical *MYD88* hotspot is enriched in a different LymphGen class than other MYD88 missense mutations. What other nuances could we be missing?

This tutorial shows how to use `annotate_curated_drivers()` from GAMBLR.utils
to add sub-gene categories to your MAF, and how to use those categories for
visualisation and statistical comparison with the companion functions
`prettyForestPlot()` (GAMBLR.viz), `test_feature_associations()` (GAMBLR.utils),
and `plot_feature_associations()` (GAMBLR.viz).

```{r load_packages}
#| message: false

library(GAMBLR.open)
library(GAMBLR.utils)
library(GAMBLR.viz)
library(tidyverse)
```

## Obtain metadata and mutations

We will compare FL and DLBCL samples across a set of recurrently mutated genes.

```{r metadata}
metadata <- get_gambl_metadata() %>%
    filter(pathology %in% c("FL", "DLBCL")) %>%
    GAMBLR.helpers::check_and_clean_metadata(duplicate_action = "keep_first")
```

```{r maf}
maf <- get_all_coding_ssm(metadata)
```

## Annotate with driver categories

`annotate_curated_drivers()` appends four new columns to every row of the MAF
for the genes you specify:

| Column | What it contains |
|---|---|
| `hot_spot` | logical; TRUE for curated hotspot mutations |
| `hotspot_alias` | the specific hotspot name (e.g. `EZH2_Y641`), NA otherwise |
| `mutation_alias` | finest-resolution label: named hotspot, structural region, or broad class (e.g. `TP53_trunc`, `TP53_R248`) |
| `driver_alias` | two-category label per gene: `<GENE>_driver` or `<GENE>_other` |

```{r annotate}
genes <- c("EZH2", "CREBBP", "TP53", "MYD88", "NOTCH1", "NOTCH2")

maf <- annotate_curated_drivers(
    maf_data        = maf,
    genes_of_interest = genes
)
```

Let's look at what was added for TP53:

```{r tp53_peek}
maf %>%
    filter(Hugo_Symbol == "TP53") %>%
    select(Tumor_Sample_Barcode, Variant_Classification,
           HGVSp_Short, hot_spot, mutation_alias, hotspot_alias, driver_alias) %>%
    head(20)
```

Notice that `hotspot_alias` is populated only for the strictest recurrent
hotspot codons and is NA for all other rows — including truncating mutations.
`mutation_alias` is more permissive: it labels truncations (`TP53_trunc`),
structural-domain categories, and hotspot codons, leaving very few rows
unlabelled. `driver_alias` collapses everything to just `TP53_driver` or
`TP53_other`.

And a quick frequency count of the alias categories:

```{r tp53_counts}
maf %>%
    filter(Hugo_Symbol == "TP53") %>%
    count(mutation_alias, sort = TRUE)
```

TP53 driver mutations are dominated by truncations (`TP53_trunc`) alongside
a handful of specific structural-domain and hotspot categories. Because
`hotspot_alias` is NA for truncating mutations, genes like TP53 — whose
driver biology is dominated by truncations — will have few or no
`hotspot_alias` entries. This distinction matters when choosing how to group
features for analysis (see below).

## Visualising associations with prettyForestPlot

`prettyForestPlot()` now accepts a `maf_column` argument that controls the
granularity of the features shown. Compare the three levels:

### Hugo_Symbol — one bar per gene (baseline)

```{r forest_gene}
prettyForestPlot(
    maf              = maf,
    metadata         = metadata,
    comparison_column = "pathology",
    comparison_values = c("FL", "DLBCL"),
    genes            = genes,
    maf_column       = "Hugo_Symbol"
)$arranged
```

### mutation_alias — finest resolution

Each curated region or hotspot codon becomes its own feature, revealing
heterogeneity within a gene. Here TP53 splits into truncation, hotspot
codon, and structural-loop categories; EZH2_Y641 stands apart from other
EZH2 mutations.

```{r forest_mutation_alias}
prettyForestPlot(
    maf              = maf,
    metadata         = metadata,
    comparison_column = "pathology",
    comparison_values = c("FL", "DLBCL"),
    genes            = genes,
    maf_column       = "mutation_alias"
)$arranged
```

This level of granularity is useful for discovery, but it comes at a cost:
splitting mutations into many fine-grained categories means each category is
supported by fewer samples. Rare categories may not be testable at all, and
the multiple-testing burden grows with the number of features. If your goal is
to ask whether a gene is differentially mutated between groups — rather than
which specific mutation class drives the difference — a coarser grouping is
usually more powerful and easier to interpret. The next two levels of
`maf_column` offer that tradeoff.

### driver_alias — binary driver classification

Collapses to just two categories per gene (`<GENE>_driver` vs `<GENE>_other`),
useful when you only want to distinguish curated driver events from background
mutations without further sub-classification.

```{r forest_driver_alias}
prettyForestPlot(
    maf              = maf,
    metadata         = metadata,
    comparison_column = "pathology",
    comparison_values = c("FL", "DLBCL"),
    genes            = genes,
    maf_column       = "driver_alias"
)$arranged
```

### hotspot_alias — strictly recurrent hotspot codons only

Only the most recurrent named hotspot codons appear here. Truncating mutations
and broad LOF regions do not generate a `hotspot_alias` entry. For TP53, whose
driver biology is dominated by truncations, very few rows appear — the
per-gene fallback mechanism in `test_feature_associations()` below handles
this case automatically.

```{r forest_hotspot_alias}
prettyForestPlot(
    maf              = maf,
    metadata         = metadata,
    comparison_column = "pathology",
    comparison_values = c("FL", "DLBCL"),
    genes            = genes,
    maf_column       = "hotspot_alias"
)$arranged
```

## Statistical testing with test_feature_associations

`prettyForestPlot()` computes statistics internally but does not expose them.
`test_feature_associations()` performs the same per-feature Fisher's exact
test and returns a tidy tibble you can inspect, filter, or feed directly into
`plot_feature_associations()`.

```{r test}
results <- test_feature_associations(
    maf              = maf,
    metadata         = metadata,
    comparison_column = "pathology",
    comparison_values = c("FL", "DLBCL"),
    genes            = genes,
    maf_column       = "mutation_alias"
)

results %>% arrange(q_value)
```

The output contains odds ratios, confidence intervals, p-values, and
Benjamini–Hochberg-adjusted q-values. The `n_mutated_<group>` and
`n_total_<group>` columns carry per-group counts and effective denominators
(only samples present in both the MAF and the metadata by default; set
`restrict_to_maf = FALSE` to include all metadata samples).

## Plotting test results with plot_feature_associations

The results tibble feeds directly into `plot_feature_associations()`. Group
sizes are carried in the `n_total_*` columns so no additional arguments are
needed for the bar plot.

```{r plot}
plots <- plot_feature_associations(
    results         = results,
    max_q           = 0.1,
    comparison_name = "FL vs DLBCL"
)

plots$arranged
```

You can filter to the most significant features and inspect the underlying
data before plotting:

```{r plot_filtered}
plots$results %>%
    select(gene, feature, OR, conf_low, conf_high, q_value,
           starts_with("n_mutated_"), starts_with("n_total_"))
```

## Advanced: controlling per-gene feature resolution

The examples above apply the same `maf_column` uniformly across all genes.
The two subsections below show how to handle genes that don't fit neatly into
a single resolution level.

### Per-gene fallback for sparse features

Some genes have no testable `hotspot_alias` entries (e.g. TP53, which is
mutated mainly through truncations that don't receive a hotspot alias). The
`fallback_column` parameter handles this gracefully: when a gene has no
features meeting `min_samples` under `maf_column`, it retests that gene using
`fallback_column` instead.

```{r test_fallback}
results_fb <- test_feature_associations(
    maf              = maf,
    metadata         = metadata,
    comparison_column = "pathology",
    comparison_values = c("FL", "DLBCL"),
    genes            = genes,
    maf_column       = "hotspot_alias",
    fallback_column  = "mutation_alias",
    verbose          = TRUE
)

# which genes used the fallback?
results_fb %>% filter(used_fallback)
```

### Manually controlling which genes use fine-grained features

The automatic fallback triggers when a gene has no testable entries under
`maf_column`. But you may want to force the fallback for a specific gene
regardless — for example, you may decide that TP53's truncation-dominated
mutation spectrum is not meaningfully split by `mutation_alias` in your cohort
and you'd prefer it to be represented by `driver_alias` categories instead,
while keeping the fine-grained `mutation_alias` resolution for all other genes.

The simplest way to achieve this is to set `mutation_alias` to `NA` for the
genes you want to exempt before calling `test_feature_associations()`. The
fallback mechanism will then take over for those genes automatically.

```{r manual_fallback}
maf_custom <- maf %>%
    mutate(mutation_alias = ifelse(Hugo_Symbol == "TP53", NA, mutation_alias))

results_custom <- test_feature_associations(
    maf              = maf_custom,
    metadata         = metadata,
    comparison_column = "pathology",
    comparison_values = c("FL", "DLBCL"),
    genes            = genes,
    maf_column       = "mutation_alias",
    fallback_column  = "driver_alias",
    verbose          = TRUE
)

# confirm TP53 fell back and others did not
results_custom %>%
    distinct(gene, used_fallback) %>%
    arrange(gene)
```

```{r plot_custom}
plot_feature_associations(
    results         = results_custom,
    max_q           = 0.1,
    comparison_name = "FL vs DLBCL"
)$arranged
```

This pattern generalises to any set of genes: just `NA` out `mutation_alias`
(or whichever `maf_column` you are using) for the genes you want treated more
coarsely, and let `fallback_column` handle them.

*Happy GAMBLing!*

```
  /$$$$$$     /$$$$$$    /$$      /$$   /$$$$$$$    /$$        .:::::::
 /$$__  $$   /$$__  $$  | $$$    /$$$  | $$__  $$  | $$        .::    .::
| $$  \__/  | $$  \ $$  | $$$$  /$$$$  | $$  \ $$  | $$        .::    .::
| $$ /$$$$  | $$$$$$$$  | $$ $$/$$ $$  | $$$$$$$   | $$   <-   .: .::
| $$|_  $$  | $$__  $$  | $$  $$$| $$  | $$__  $$  | $$        .::  .::
| $$  \ $$  | $$  | $$  | $$\  $ | $$  | $$  \ $$  | $$        .::    .::
|  $$$$$$/  | $$  | $$  | $$ \/  | $$  | $$$$$$$/  | $$$$$$$$  .::      .::
 \______/   |__/  |__/  |__/     |__/  |_______/   |________/
 ~GENOMIC~~~~~~~~~~~~~OF~~~~~~~~~~~~~~~~~B-CELL~~~~~~~~~~~~~~~~~~IN~~~~~~
 ~~~~~~~~~~~~ANALYSIS~~~~~~MATURE~~~~~~~~~~~~~~~~~~~LYMPHOMAS~~~~~~~~~~R~
```
