---
title: "Why GAMBLR.open?"
vignette: >
  %\VignetteIndexEntry{quarto vignettes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: html
warning: false
execute:
    cache: false
---

## Open data access, simplified!

GAMBLR.open is designed to make open-access genomic data of mature B-cell lymphomas more accessible and easier to analyze. Whether you're a researcher, data scientist, or student, this package provides a seamless interface for querying, retrieving, and preprocessing genomic data, so you can focus on analysis rather than data wrangling.

GAMBLR.open is part of a larger family of R tools for the analysis of lymphoma
genomic data (*the GAMBLRverse*, if you will). GAMBLR began as a toolkit for analysts participating
in the **G**enomic **A**nalysis of **M**ature **B**-cell **L**ymphomas project, 
led by the [Morin Lab](https://morinlab.github.io/) at BC Cancer and Simon Fraser
University.

```
  /$$$$$$     /$$$$$$    /$$      /$$   /$$$$$$$    /$$        .:::::::     
 /$$__  $$   /$$__  $$  | $$$    /$$$  | $$__  $$  | $$        .::    .::   
| $$  \__/  | $$  \ $$  | $$$$  /$$$$  | $$  \ $$  | $$        .::    .::   
| $$ /$$$$  | $$$$$$$$  | $$ $$/$$ $$  | $$$$$$$   | $$   <-   .: .::       
| $$|_  $$  | $$__  $$  | $$  $$$| $$  | $$__  $$  | $$        .::  .::     
| $$  \ $$  | $$  | $$  | $$\  $ | $$  | $$  \ $$  | $$        .::    .::   
|  $$$$$$/  | $$  | $$  | $$ \/  | $$  | $$$$$$$/  | $$$$$$$$  .::      .:: 
 \______/   |__/  |__/  |__/     |__/  |_______/   |________/ 
 ~GENOMIC~~~~~~~~~~~~~OF~~~~~~~~~~~~~~~~~B-CELL~~~~~~~~~~~~~~~~~~IN~~~~~~ 
 ~~~~~~~~~~~~ANALYSIS~~~~~~MATURE~~~~~~~~~~~~~~~~~~~LYMPHOMAS~~~~~~~~~~R~ 
```

## Who Is It For?

The GAMBLR.open primarily caters to:

- researchers who need rapid access to standardized genomic datasets.
- data scientists and bioinformaticians looking for structured and easily queryable cancer genomics data.
- students who want to explore real-world genomic data in a learning environment.

## What does GAMBLR.open offer?

Some of the advantages of GAMBLR.open are:

- Access to preprocessed genomic data: retrieve genomic datasets without worrying about complex file handling.
- Standardized data queries: easily search and filter genomic data using a consistent and intuitive API.
- Integration of GAMBLR ecosystem: work seamlessly with the broader [GAMBLR suite](./concepts/GAMBLR_family.html) of packages to design analytical workflows.
- Reproducibility and transparency: leverage a well-documented [pipelines](https://github.com/LCR-BCCRC/lcr-modules) to ensure consistent and transparent data retrieval.


## The GAMBLR.open family includes:

### Overview

* **GAMBLR.open** - this package.

* **GAMBLR.data** - a "storage unit" with various types of bundled data including curated resources.

* **GAMBLR.viz** - A collection of functions to help generate visual representations of lymphoma genomic data. 

* **GAMBLR.helpers** - a set of low-level functions for data operation. 

* **GAMBLR.utils** - higher level set of somewhat generic functions to facilitate some common cancer genomic data analysis tasks. 

* **GAMBLR.predict** - a collection of machine learning algorithms and functions to pre-format inputs for these models. Contains classifiers of [Burkitt](https://doi.org/10.1182/blood.2022016534) and [Follicular](https://doi.org/10.1182/blood.2022018719) lymphomas originally published, as well as reproduced of DLBCL classification by the groupings of [Chapuy et al](https://doi.org/10.1038/s41591-018-0016-8), [Lacy et al](https://doi.org/10.1182/blood.2019003535), and [Runge et al](https://doi.org/10.1111/bjh.17132).

### Code

* [**GAMBLR.open**](https://github.com/morinlab/GAMBLR.open)

* [**GAMBLR.data**](https://github.com/morinlab/GAMBLR.data)

* [**GAMBLR.viz**](https://github.com/morinlab/GAMBLR.viz)

* [**GAMBLR.helpers**](https://github.com/morinlab/GAMBLR.helpers)

* [**GAMBLR.utils**](https://github.com/morinlab/GAMBLR.utils)

* [**GAMBLR.predict**](https://github.com/morinlab/GAMBLR.predict)

### Documentation and tutorials

* [**GAMBLR.open**](https://morinlab.github.io/GAMBLR.open)

* [**GAMBLR.data**](https://morinlab.github.io/GAMBLR.data/resources/bundled_data.html) 

* [**GAMBLR.viz**](https://morinlab.github.io/GAMBLR.viz)



## Dependencies

This diagram illustrates in a simplified way the dependency relationship of
different packages in the GAMBLR family:


```{mermaid}
%%{
    init : {
        "flowchart" : { "curve" : "monotoneX" },
        'themeVariables': {'fontSize': '24px'}
    }

}%%
flowchart LR
  A("GAMBLR.data") ==> B("GAMBLR.helpers")
  B ==> C("GAMBLR.utils")
  B ==> D("GAMBLR.predict")
  D ==> E("GAMBLR.open")
  C ==> F("GAMBLR.viz")
  F ==> E
  classDef A,B font-size:18pt;
  style E fill:#6aa84f,stroke:#333,stroke-width:4,font-size:20px
```


## Quickstart

The quick and easy way to get started is to make sure the `devtools` dependency is installed, then install the GAMBLR.open and run the simplest functions to confirm successful setup:

```{r}
#| eval: false
# Verify devtools is installed
if (!require("devtools")) install.packages("devtools")

# Install GAMBLR.open
devtools::install_github(
    "morinlab/GAMBLR.open",
    repos = BiocManager::repositories()
)
```

We recommend restarting your R session after the installation.

```{r}
#| eval: false
library(GAMBLR.open)

# Confirm successful setup
metadata <- get_gambl_metadata()
maf <- get_coding_ssm()
```

## Installation for developers

The easiest way to obtain and contribute to GAMBLR.open is to do this via cloning the repository

```{bash}
#| eval: false
cd
git clone git@github.com:morinlab/GAMBLR.open.git
```
In your R editor of choice (which is __**hopefully**__ [VS Code](https://code.visualstudio.com/)), set your working directory to the directory you just cloned the repo into.

```{bash}
#| eval: false
setwd("~/GAMBLR.open")
```

Install the package in R by running the following command (requires the
`devtools` package):

```{r}
#| eval: false
devtools::install()
```

After applying your modifications to the code, use the following command to
quickly test your changes without directly installing the package (requires the
`devtools` dependency):

```{r}
#| eval: false
devtools::load_all()
```

## Community

GAMBLR.open is a free open-source package, but the Master branch is protected.
We welcome contributions (pull request, bug report, feature request, PR review)
from all levels of users. All commits must be submitted via pull request on a
branch. Please refer to the GitHub [documentation](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-a-pull-request) for details on how to do pull request.
